---
title: "DAGs"
author: "Ilya Moskalenko"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
  output:
  pdf_document:
    toc: yes
---


*******

\newpage

# Part 0: Introduction

## 0.1: Resources

**ggdag**
https://cran.r-project.org/web/packages/ggdag/readme/README.html#

**DiagrammeR**
https://www.graphviz.org/doc/info/shapes.html
https://rich-iannone.github.io/DiagrammeR/index.html

## 0.2: Package Installation

```{r}
# install.packages("ggdag")
# install.packages("DiagrammeR")
# install.packages("dagitty")
```

## 0.3: Loading Libraries

```{r, message=FALSE}
library(ggdag)
library(dagitty)
library(tidyverse)
library(DiagrammeR)
```

## 0.4: Creating one layer for beautification 

```{r}
# ggdag uses the same syntax as ggplot
beautification <- theme_minimal() + 
  theme(panel.grid.major = element_blank(),  # Remove major grid lines
        panel.grid.minor = element_blank(),  # Remove minor grid lines
        axis.text = element_blank(),         # Remove axis labels
        axis.ticks = element_blank(),        # Remove axis tick marks
        axis.title = element_blank())        # Remove axis titles
```

\newpage

# Part 1: ggdag

## 1.1 DAG Setup: Mapping Nodes Positions, Mapping Arrows, and Labelling Nodes

```{r}
# Define the x and y coordinates for nodes in the DAG (Directed Acyclic Graph)
node_positions <- list(
  x = c(A = 0, W1 = 0, W2 = 1, D = 0.5,  Y = 1),  # X-coordinates for nodes
  y = c(A = 0, W1 = 1, W2 = 1, D = -1,  Y = 0)   # Y-coordinates for nodes
)

# Create a DAG (Directed Acyclic Graph) using the dagify function
dag <- dagify(
  A ~ W1 + W2,           # Define the edges in the DAG
  Y ~ A + W1 + W2,
  D ~ A + Y,
  labels = c(            # Assign labels to nodes
    A = "Smoking\n (Exposure)",    # Label for node A
    Y = "Obesity\n (Outcome)",     # Label for node Y
    D = "Diabetes",                # Label for node D
    W1 = "SES",                    # Label for node W1
    W2 = "Education"               # Label for node W2
  ),
  coords = node_positions,  # Use the specified node positions
  exposure = "A",           # Specify the exposure variable
  outcome = "Y"            # Specify the outcome variable
)
```

## 1.2: Visualizing simple DAG

```{r}
dag %>%
  ggdag() # Like ggplot only for DAGs
```

## 1.3: Visualizing simple DAG + Beautification

```{r}
dag %>%
  ggdag() + 
  beautification # Adding Beautification
```

## 1.4: Visualizing the adjustment set

```{r}
dag %>%
  ggdag_adjustment_set() + # Showing Adjustment sets 
  beautification
```

## 1.5: Visualizing pathways between nodes with shadows

```{r}
dag %>% 
  ggdag_paths(from="A", to= "Y", # Specifying starting and ending nodes
              adjust_for = "W2", # Adjusting for a node
              shadow = TRUE) + # Showing a shadow for the other nodes rather than removing them (Best Practice)
  beautification
```

## 1.6: Visualizing pathways between nodes without shadows

```{r}
dag %>% 
  ggdag_paths(from="A", to= "Y", # Specifying starting and ending nodes
              adjust_for = "W2", # Adjusting for a node
              shadow = FALSE) + # Only showing pathways (NOT best practice)
  beautification
```

## 1.7: Visualizing with neat labels

```{r}
dag %>% 
  ggdag(text = FALSE, # Do NOT use A, Y, W1, W2, D
        use_labels = "label") + # Use the corresponding labels 
  beautification
```


*******

\newpage

# Part 2: dagitty + ggdag

A good **use case** for using the dagitty website is when you are performing an **exploratory analysis** and are playing around with a lot of variables.

## 2.1: Navigate to Dagitty Website, Create a DAG, Copy the Code

Dagitty Website: https://dagitty.net
  
## 2.2: Paste the Code into Code Chunk + Wrap in dagitty() function

Make sure to wrap the **dagitty()** function around the code copied from Dagitty website.

```{r}
dag <- dagitty('dag {
A [exposure,pos="-1.642,0.826"] 
D [pos="-0.502,1.607"]
W1 [pos="-1.666,-0.450"]
W2 [pos="0.656,-0.465"]
Y [outcome,pos="0.662,0.865"]
A -> D
A -> Y
W1 -> A
W1 -> Y
W2 -> A
W2 -> Y
Y -> D
}
')

dag %>% 
  ggdag() + beautification
```


### 2.2.1: Paste the Code into Code Chunk + Wrap in dagitty() function with Comments

'dag {
A [exposure,pos="-1.372,1.120"]    # Node A with exposure and position
D [pos="-0.339,1.675"]             # Node D with position
W1 [pos="-1.369,0.288"]            # Node W1 with position
W2 [pos="0.661,0.288"]             # Node W2 with position
Y [outcome,pos="0.668,1.149"]      # Node Y as an outcome variable with position
A -> D                             # Directed edge from A to D
A -> Y                             # Directed edge from A to Y
W1 -> A                            # Directed edge from W1 to A
W1 -> Y                            # Directed edge from W1 to Y
W2 -> A                            # Directed edge from W2 to A
W2 -> Y                            # Directed edge from W2 to Y
Y -> D                             # Directed edge from Y to D
}
'

## 2.3: Manually Adjusting Coordinates for Visual Symmetry

```{r}
dag <- dagitty('dag {
D [pos="0,1.5"]
W1 [pos="1,0.5"]
Y [outcome,pos="1,1"]
W2 [pos="-1,0.5"]
A [exposure,pos="-1,1"]
A -> D
A -> Y
W1 -> A
W1 -> Y
W2 -> A
W2 -> Y
Y -> D
}
')

dag %>% 
  ggdag() +
  beautification
```

\newpage

*******

# Part 3: DiagrammeR

## 3.1: Simple DAG

```{r}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2] 
  node [shape = plaintext]
    A 
    Y 
    D 
    W1
    W2
  edge [minlen = 2]
    A->Y
    A->D
    Y->D
    W1->A
    W1->Y
    W2->A
    W2->Y
    
  { rank = same; A; Y }
}
")
```

### 3.1: Simple DAG with Comments

DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]       # Set the vertical distance between ranks in the graph
  node [shape = plaintext]    # Set the shape of nodes as plaintext (for custom labels)
    A                         # Define node A
    Y                         # Define node Y
    D                         # Define node D
    W1                        # Define node W1
    W2                        # Define node W2
  edge [minlen = 2]           # Set the minimum length of edges to 2 units
    A->Y                      # Create an edge from A to Y
    A->D                      # Create an edge from A to D
    Y->D                      # Create an edge from Y to D
    W1->A                     # Create an edge from W1 to A
    W1->Y                     # Create an edge from W1 to Y
    W2->A                     # Create an edge from W2 to A
    W2->Y                     # Create an edge from W2 to Y
    
  { rank = same; A; Y }       # Specify that nodes A and Y should be on the same rank
}
")

## 3.2: Adding Node Labels

```{r}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext]
    A [label = 'Smoking']
    Y [label = 'Obesity']
    D [label = 'Diabetes']
    W1 [label = 'SES']
    W2 [label = 'Education']
  edge [minlen = 2]
    A->Y
    A->D
    Y->D
    W1->A
    W1->Y
    W2->A
    W2->Y
    
  { rank = same; A; Y }
}
")
```

## 3.3: Modifying Node Spacing

```{r}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.4]
  node [shape = plaintext]
    A [label = 'Smoking']
    Y [label = 'Obesity']
    D [label = 'Diabetes']
    W1 [label = 'SES']
    W2 [label = 'Education']
  edge [minlen = 2]
    A->Y
    A->D
    Y->D
    W1->A
    W1->Y
    W2->A
    W2->Y
    
  { rank = same; A; Y }
}
")
```

## 3.4: Modifying Edge Length

```{r}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.4]
  node [shape = plaintext]
    A [label = 'Smoking']
    Y [label = 'Obesity']
    D [label = 'Diabetes']
    W1 [label = 'SES']
    W2 [label = 'Education']
  edge [minlen = 4]
    A->Y
    A->D
    Y->D
    W1->A
    W1->Y
    W2->A
    W2->Y
    
  { rank = same; A; Y }
}
")
```

## 3.5: Adding Boxes to Indicate Adjusted Variables

```{r}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.4]
  node [shape = plaintext]
    A [label = 'Smoking']
    Y [label = 'Obesity']
    D [label = 'Diabetes']
    W1 [label = 'SES', shape = box]
    W2 [label = 'Education', shape = box]
  edge [minlen = 4]
    A->Y
    A->D
    Y->D
    W1->A
    W1->Y
    W2->A
    W2->Y
    
  { rank = same; A; Y }
}
")
```

