---
title: "Leaflet Module 1: Introduction"
author: "Ilya Moskalenko"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    df_print: paged
---


\newpage

***************************************************

# Part 1: Intro to Leaflet

Hi All, welcome to GIS using the Leaflet R library!

## 1.1: Resources

Info on Admin Boundaries:

- [Link](https://en.wikipedia.org/wiki/List_of_administrative_divisions_by_country)

Leaflet GitHub Repo:

- [Link](https://rstudio.github.io/leaflet/)

Extra Leaflet Practice:

- [Link to Geeks for Geeks](https://www.geeksforgeeks.org/leaflet-package-in-r/)

- [Link to Earth Data Science](https://www.earthdatascience.org/courses/earth-analytics/get-data-using-apis/leaflet-r/)

## 1.2: Libraries

```{r libraries, message=FALSE}
library(tidyverse)
#install.packages("leaflet")
library(leaflet) # Interactive mapping library
library(sp) # Spatial data objects library
library(janitor) # Clean names library
library(viridis) # Accessible color palette library
```

***************************************************

\newpage

# Part 2: Data Preprocessing

## 2.1: Read in Data

For this module, we will use an open source Hospital data set from ArcGIS Online. Specifically, this dataset provides the locations of all the Hospitals in the United States.

For more information you can navigate to the data source using the following link:

- [Link](https://hifld-geoplatform.opendata.arcgis.com/datasets/hospitals/explore?location=41.782325%2C-115.894111%2C4.12&showTable=true)

```{r data import}
# reading in 'Hospital' df and filtering where state is California
df_hospitals <- read_csv('Hospitals.csv') %>%
  filter(STATE != 'PR') %>% # Filtering out Puerto Rico
  clean_names() # From janitor package (columns lowercased & replaces punctuation and spacing with "_")
```
## 2.2: Look at Data

The data set has 34 features (another word for columns) and therefore has a lot of potential for rich analyses and visualizations. For the purposes of this module, we will only be leveraging 7 features:

- **latitude**: latitude of hospital

- **longitude**: longitude of hospital

- **name**: name of hospital

- **address**: address of hospital

- **beds**: bed capacity of hospital

- **telephone**: telephone of hospital

- **owner**: ownership type of the hospital (private, non-profit, government)

```{r}
head(df_hospitals)

summary(df_hospitals)
```

As we can see, one of the features that we are using **beds** has negative values which is odd for something that should not be a negative value. Specifically, -999 is often used to signify a **missing value**. Therefore, I will **filter** out all negative values from this feature (I will actually filter for beds > 200 so the visualizations are a bit less cluttered). 

We will also be using the **owner** column and I want to clean up those values a bit. I will leverage the **stringr** library (part of tidyverse) to abbreviate "GOVERNMENT" and also transform the values to title case (All word lowercased with first letter of each word being uppercased).   

This **highlights** the importance at taking a look at your data (especially the features that are of interest) before processing it.

```{r}
df_hospitals <- df_hospitals %>% 
  filter(beds>200) %>%
  mutate(owner= str_to_title(str_replace(owner,"GOVERNMENT -", "GOV")))

head(df_hospitals)
```

## 2.3: Create Spatial Points Data Frame (SPDF)

To build a map leveraging **Leaflet**, we need to transform our data into a **Spatial Points Data Frame (SPDF)**. An **SPDF** is a data structure commonly used in geographical information systems (GIS) and spatial analyses. It is a specific data format that combines the attributes of a traditional data frame or table with spatial information, such as coordinates (latitude and longitude) to represent point locations on the Earth's surface. 

```{r SPDF}
# Create a new SpatialPointsDataFrame object named 'spdf_hospitals' to represent geographical data
spdf_hospitals <- SpatialPointsDataFrame(
  # Specify the coordinates for the points using the 'longitude' and 'latitude' columns from 'df_hospitals'
  # Reminder, we can subset columns in a data frame using this format: df[,c('a','b')]
  coords = df_hospitals[,c('longitude','latitude')], 
  # Include data attributes associated with each point using selected columns from 'df_hospitals'
  data = df_hospitals[,c('name','address','beds','telephone', 'owner')],
  # Define the projection information for the data, using EPSG 4326
  
  # NOTE: this line of code was updated from the video module to reflect package retirement
  proj4string = CRS(SRS_string = "EPSG:4326") # CRS is "Coordinate Reference System"
)
```

***************************************************

# Part 3: Mapping

## 3.1: Initial Map Layer

```{r}
# Create a Leaflet map object
map <- leaflet() %>% # Similar to intial ggplot() layer
  setView(lng = -96.25, lat = 39.50, zoom = 4) # Setting the Zoom to the US 
```

## 3.2: Map Themes

Here are a few preset themes using **addProviderTiles()** that you can use to determine the topography of your map and you can navigate to the following link for more themes!

- [Link](https://leaflet-extras.github.io/leaflet-providers/preview/)

```{r}
# Default basic theme
map %>% addTiles()

# Use a theme from the Leaflet Providers plug-in - CartoDB Positron
map %>% addProviderTiles(providers$CartoDB.Positron)

# Use the "Stamen.Toner" theme for the map
map %>% addProviderTiles("Stamen.Toner")

# Replace or add a new comment for the next line: Use the "Esri.NatGeoWorldMap" theme for the map
map %>% addProviderTiles("Esri.NatGeoWorldMap")

# Specify a custom map tile URL template as an argument to addTiles
map %>% addTiles("https://tiles.stadiamaps.com/tiles/alidade_smooth_dark/{z}/{x}/{y}{r}.png")

# Explore more built-in Provider themes in R - in this case, CartoDB Voyager
map %>% addProviderTiles(providers$CartoDB.Voyager)
```

## 3.3: Markers

**addCircleMarkers()** is the function that allows us to add the data points from the SPDFs that we created onto the map.

```{r}
map %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
# -------------------------------------------------------------------------  
  # Add circle markers for hospital data
  addCircleMarkers(
    data = spdf_hospitals,
    color = 'white',    # Set the outline color to white
    weight = 1.5,         # Set the weight of the outline
    fillColor = 'blue',  # Set the fill color to blue
    fillOpacity = 0.7,    # Set fill opacity for the circle markers
    radius = 2.5
  ) 
```

## 3.4: Modify Marker Information 

### 3.4.1: Popups

**Popups** are clickable markers that present information regarding that spatial data point.

```{r}
map %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = spdf_hospitals,
    color = 'white',    
    weight = 1.5,       
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
# -------------------------------------------------------------------------    
    # Define the popup content for each circle marker
    popup = paste(      
      'Name:', spdf_hospitals$name, '<br>',
      'Address:', spdf_hospitals$address, '<br>',
      'Beds:', spdf_hospitals$beds, '<br>',
      'Phone:', spdf_hospitals$telephone, '<br>',
      'Owner:', spdf_hospitals$owner, '<br>')
# -------------------------------------------------------------------------
  ) 
```

### 3.4.2: Labels

**Labels** are hover markers that present information regarding that spatial data point.

```{r}
map %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = spdf_hospitals,
    color = 'white',    
    weight = 1.5,       
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
# -------------------------------------------------------------------------        
    label = paste(      # Define the popup content for each circle marker
      'Hospital:', spdf_hospitals$name),
    labelOptions = labelOptions(  
        style = list(
                    "font-family" = "serif",  # Font
                    "font-family" = "serif",  # Font color
                    "font-style" = "italic",  # Italicizing font
                    "font-size" = "12px")))   # Font size
```

### 3.4.3: Cluster

```{r}
map %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = spdf_hospitals,
    color = 'white',    
    weight = 1.5,       
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
# --------------------------------------------------------------------    
    clusterOptions = markerClusterOptions(), # Add this line of code
# --------------------------------------------------------------------
) 
```

## 3.5: Add Legend

```{r}
map %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = spdf_hospitals,
    color = 'white',    
    weight = 1.5,       
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(      
      'Name:', spdf_hospitals$name, '<br>',
      'Address:', spdf_hospitals$address, '<br>',
      'Beds:', spdf_hospitals$beds, '<br>',
      'Phone:', spdf_hospitals$telephone, '<br>',
      'Owner:', spdf_hospitals$owner, '<br>')) %>%
# -------------------------------------------------------------------------  
  # Add a legend to the map
  addLegend(
    position = 'topright',   # Position the legend in the top-right corner
    title = 'Marker Colors', # Title for the legend
    labels = c('Hospital'), # Labels for the legend
    colors = c('blue'),      # Corresponding colors for the labels
    opacity = 1             # Set the opacity of the legend
  )
```

***************************************************

\newpage

# Part 4: Color Modifications

## 4.1: Continuous Variables

```{r}
# --------------------------------------------------------------------------  
# Define a color palette based on the "beds" column
color_palette <- colorNumeric(
  palette = viridis(7),  # Choose a color palette (e.g., viridis)
  domain = spdf_hospitals$beds)  # Specify the data column for color mapping
# --------------------------------------------------------------------------  
map %>%
 addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    data = spdf_hospitals,
# --------------------------------------------------------------------------    
    color = color_palette(spdf_hospitals$beds),  # Use the color palette
# --------------------------------------------------------------------------  
    weight = 1,
    fillOpacity = 0.7,
    radius = 1.5,
    popup = paste(
      'Name:', spdf_hospitals$name, '<br>',
      'Address:', spdf_hospitals$address, '<br>',
      'Beds:', spdf_hospitals$beds, '<br>',
      'Phone:', spdf_hospitals$telephone, '<br>'
    )
  ) %>%
  addLegend(
    position = "topright",
# --------------------------------------------------------------------------  
    pal = color_palette,              # Updating the legend to reflect color
    values = spdf_hospitals$beds,   
    title = "Beds",                
# --------------------------------------------------------------------------  
    opacity = 1
  )
```

## 4.2: Categorical Variables

```{r}
color_palette <- colorFactor(
  palette = viridis(7),  
# --------------------------------------------------------------------------  
  domain = spdf_hospitals$owner  # Specify the data column for color mapping
# --------------------------------------------------------------------------  
)

map %>%
 addProviderTiles(providers$CartoDB.Positron) %>%
 addCircleMarkers(
   data = spdf_hospitals,
# --------------------------------------------------------------------------  
   color = ~color_palette(spdf_hospitals$owner),  # Use the color palette
# --------------------------------------------------------------------------  
   weight = 1,
   fillOpacity = 0.7,
   radius = 1.5,
   popup = paste(
     'Name:', spdf_hospitals$name, '<br>',
     'Address:', spdf_hospitals$address, '<br>',
     'Owner:', spdf_hospitals$owner, '<br>',
     'Phone:', spdf_hospitals$telephone, '<br>'
   )
 ) %>%
 addLegend(
   position = "topright",
   pal = color_palette,  
# --------------------------------------------------------------------------  
   values = spdf_hospitals$owner, # Updating the legend to reflect color
   title = "Owner",
# -------------------------------------------------------------------------- 
   opacity = 1
 )
```


**Leaflet Module 2** will pull on concepts from this module and teach how to implement **Layer Control** a neat interactive mapping feature!
