---
title: "Leaflet Module 2: Layer Control"
author: "Ilya Moskalenko"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    df_print: paged
---


***************************************************

# Part 1: Loading Libraries and Reading in Data 


```{r libraries, message=FALSE}
library(tidyverse)
library(leaflet)
library(sp)
library(janitor)
```

For this activity we will be **Mapping Prominent Hospital Systems in California using Leaflet**.

As done in **Leaflet Module 1**, we will read in the data set. Since we are only mapping the prominent hospital systems in California, we can filter the data set during our read in step for only hospitals that are located within California. 

```{r data import}
# reading in 'Hospital' df and filtering where state is California
df_hospitals <- read_csv('Hospitals.csv') %>%
  filter(STATE == 'CA') %>% # Filtering out Puerto Rico
  clean_names() # From janitor package (columns lowercased & replaces punctuation and spacing with "_")
```
# Part 2: Creating Health Network Specific Data Frames

As part of this Leaflet module, I will be teaching **Layer Control** which allows for a user to select which features appear on the map. Since we are interested in creating a map of the prominent health networks in California, we will first need to create data frames that are specific to each health network. 

```{r filtering data}
# UCLA hospitals df
df_ucla <- df_hospitals %>%
  filter(str_detect(name, 'UCLA')) # 'str_detect()' is part of the stringr package

# UCSF hospitals df
df_ucsf <- df_hospitals %>%
  filter(str_detect(name, 'UCSF'))

# Cedars Sinai hospitals df
df_cedars <- df_hospitals %>%
  filter(str_detect(name, 'CEDARS'))

# Kaiser Permanente hospitals df
df_kaiser <- df_hospitals %>%
  filter(str_detect(name, 'KAISER'))

# UCSD hospitals df
df_ucsd <- df_hospitals %>%
  filter(str_detect(name, 'UC SAN DIEGO'))

# Stanford hospitals df
df_stanford <- df_hospitals %>%
  filter(str_detect(name, 'STANFORD'))
```

***************************************************

# Part 3: Creating Health Network Specific SPDFs

We will now create a Spatial Points Data Frame (SPDF) for each dataframe that is specific to each healthcare network using the healthcare specific data frames created in the previous step. 

If you need a refresher on what an SPDF is, please refer back to the Intro Leaflet Module.

```{r SPDF}
# Create SpatialPointsDataFrame for UCLA hospitals
spdf_ucla <- SpatialPointsDataFrame(
  coords = df_ucla[,c('longitude','latitude')],
  data = df_ucla[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video

# Create SpatialPointsDataFrame for UCSF hospitals
spdf_ucsf <- SpatialPointsDataFrame(
  coords = df_ucsf[,c('longitude','latitude')],
  data = df_ucsf[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video

# Create SpatialPointsDataFrame for Kaiser Permanente hospitals
spdf_kaiser <- SpatialPointsDataFrame(
  coords = df_kaiser[,c('longitude','latitude')],
  data = df_kaiser[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video 

# Create SpatialPointsDataFrame for Cedars Sinai hospitals
spdf_cedars <- SpatialPointsDataFrame(
  coords = df_cedars[,c('longitude','latitude')],
  data = df_cedars[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video

# Create SpatialPointsDataFrame for UCSD hospitals
spdf_ucsd <- SpatialPointsDataFrame(
  coords = df_ucsd[,c('longitude','latitude')],
  data = df_ucsd[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video

# Create SpatialPointsDataFrame for Stanford hospitals
spdf_stanford <- SpatialPointsDataFrame(
  coords = df_stanford[,c('longitude','latitude')],
  data = df_stanford[,c('name','address','beds','telephone')],
  proj4string = CRS(SRS_string = "EPSG:4326")) # NOTE: updated from the video
```

***************************************************

# Part 4: Setting Up Map Layers

```{r}
# Create a Leaflet map object
map <- leaflet() %>%
  
  # Set the base map tiles to CartoDB Positron (a light-colored map)
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # UCLA Layer
  addCircleMarkers(
    data = spdf_ucla, # Specify UCLA SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(      
      'Corporation name:', spdf_ucla$name, '<br>', # Specify "name" column in UCLA SPDF
      'address:', spdf_ucla$address, '<br>',       # Specify "address" column in UCLA SPDF
      'beds:', spdf_ucla$beds, '<br>',             # Specify "beds" column in UCLA SPDF 
      'Phone:', spdf_ucla$telephone, '<br>'        # Specify "telephone" column in UCLA SPDF 
    ),
    group="UCLA"                                   # Group is UCLA
  ) %>%
  
  # UCSF Layer
  addCircleMarkers(
    data = spdf_ucsf,# Specify UCSF SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(
      'Corporation name:', spdf_ucsf$name, '<br>', # Specify "name" column in UCSF SPDF
      'address:', spdf_ucsf$address, '<br>',       # Specify "address" column in UCSF SPDF
      'beds:', spdf_ucsf$beds, '<br>',             # Specify "beds" column in UCSF SPDF 
      'Phone:', spdf_ucsf$telephone, '<br>'        # Specify "telephone" column in UCSF SPDF 
    ),
    group="UCSF" # Group is UCSF
  ) %>%
  
  # Cedars Sinai Layer
  addCircleMarkers(
    data = spdf_cedars, # Specify Cedars SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(
      'Corporation name:', spdf_cedars$name, '<br>', # Specify "name" column in Cedars SPDF
      'address:', spdf_cedars$address, '<br>',       # Specify "address" column in Cedars SPDF
      'beds:', spdf_cedars$beds, '<br>',             # Specify "beds" column in Cedars SPDF 
      'Phone:', spdf_cedars$telephone, '<br>'        # Specify "telephone" column in Cedars SPDF 
    ),
    group= "Cedars Sinai"
  ) %>%
  
  # Kaiser Permanente Layer
  addCircleMarkers(
    data = spdf_kaiser, # Specify Kaiser SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(
      'Corporation name:', spdf_kaiser$name, '<br>', # Specify "name" column in Kaiser SPDF
      'address:', spdf_kaiser$address, '<br>',       # Specify "address" column in Kaiser SPDF
      'beds:', spdf_kaiser$beds, '<br>',             # Specify "beds" column in Kaiser SPDF 
      'Phone:', spdf_kaiser$telephone, '<br>'        # Specify "telephone" column in Kaiser SPDF
    ),
    group= "Kaiser Permanente"
  ) %>%
  
  # UCSD Layer
  addCircleMarkers(
    data = spdf_ucsd, # Specify UCSD SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(
      'Corporation name:', spdf_ucsd$name, '<br>', # Specify "name" column in UCSD SPDF
      'address:', spdf_ucsd$address, '<br>',       # Specify "address" column in UCSD SPDF
      'beds:', spdf_ucsd$beds, '<br>',             # Specify "beds" column in UCSD SPDF 
      'Phone:', spdf_ucsd$telephone, '<br>'        # Specify "telephone" column in UCSD SPDF
    ),
    group = "UCSD"
  ) %>%
  
  # Stanford Layer
  addCircleMarkers(
    data = spdf_stanford, # Specify Stanford SPDF
    color = 'white',    
    weight = 1,         
    fillColor = 'blue', 
    fillOpacity = 0.7,  
    radius = 2.5,
    popup = paste(
      'Corporation name:', spdf_stanford$name, '<br>', # Specify "name" column in Stanford SPDF
      'address:', spdf_stanford$address, '<br>',       # Specify "address" column in Stanford SPDF
      'beds:', spdf_stanford$beds, '<br>',             # Specify "beds" column in Stanford SPDF 
      'Phone:', spdf_stanford$telephone, '<br>'        # Specify "telephone" column in Stanford SPDF
    ), 
    group = "Stanford"
  ) %>%
  
  # Add a legend to the map
  addLegend(
    position = 'topright',  
    title = 'Marker Colors',
    labels = c('Hospital'), 
    colors = c('blue'),     
    opacity = 1             
  ) 
```

***************************************************

# Part 5: Implementing Layer Control

## 5.1: Layer Control Drop Down

```{r}
map %>%
# Add a layer control
  addLayersControl(
    overlayGroups = c("UCLA", "UCSF", "Cedars Sinai", 
                      "Kaiser Permanente", "UCSD", "Stanford"),   
    options = layersControlOptions(collapsed = TRUE) # Collapsing the Layer control
  )
```

## 5.2: Layer Control Default

```{r}
map %>%
# Add a layer control
  addLayersControl(
    overlayGroups = c("UCLA", "UCSF", "Cedars Sinai", 
                      "Kaiser Permanente", "UCSD", "Stanford"),   
    options = layersControlOptions(collapsed = FALSE) # Expanded by default
  )
```

As you may have noticed, this code was all very repetitive! **Leaflet Module 3: Automation with Leaflet** will cover **Automation** to optimize your code. In this module you will learn how to utilize data structures, leverage functions and loops to **Automate** your code. 
